### 2023-02-06 article.tex
Improve structure

\setcounter{chapter}{7}
\chapter{Research code is as vital paradata as the article itself}

@@

Here we define paradata as data about the data collection process.
In genetic epidemiology, the data generated are mostly the results 
of an analysis (e.g. predicting a person having a disease),
as done by computer code.
In such context, one could argue incorrectly that
it is the scientific paper that best describes which computation steps 
have taken place.
However, this has the unrealistic 
assumption that there is a perfect match between the paper and the code.
In this chapter it is argued that the source code should be supplied,
as these are the true paradata: if the paper and code disagree, it is the
code that has generated the results.
The chapter concludes by some rules how to better code to serve as paradata,
and hence allowing computational research to be truly reproducible.

@@

% Alex and Blake?
Two different researchers in genetic epideomology (more on that field later), 
write two equally good manuscripts
that involve computation.
Both manuscripts are accepted by an equally prestigous journal after peer view. 
One researcher, however, does not supply the
computer code that was used to generate the results,
where the other does.
Are the conclusion of these papers to be trusted equally?
Is this difference relevant and worth the effort?
This chapter discusses the effect of supplying the computer 
code (from now on: 'code') 
and claims that code is paradata and that this paradata 
is just as important as the paper itself.

@@

A recent paper explores the use of paradata to increase the impact of data,
stating that lack of paradata can be seen as 'a drastic constraint'
in the use of data and offer some suggestions to 
make paradata useful for data re(use):
paradata should be comprehensive, comprehensively documented
and friendly to computers \cite{huvila2022improving}.
These suggestions appear to work with code as well,
where these suggestions show up as best practices (such
as \cite{stodden2013best}), some of which
described below.
The lack of computer code is just as much 'a drastic constraint'
in the use of a scientific experiment.
The clearest effect of its absence is, most obviously,
that a computational experiment cannot be redone 
easily, yet there are wider implications, described below, as well.

@@

This paper uses genetic epidemiology as a specific example,
to illustrate that the computer code is paradata,
that this type of paradata is relevant, as it helps 
to assess the correctness of the results,
where the correctness of the results is relevant for 
science in general and healthcare in particular.
However, any field that uses computation in its experiments
could be used as an example.

@@

Going back to the example scientists in the introduction,
where one does and the other does not publish the code,
it is not common to publish the code of an experiment or analysis.
For example, in computer graphics, 
a field intimately familiar with computer code,
only 192 of 454 SIGGRAPH papers supply computer code \cite{bonneel2020code}.
Another study analysed the reproducibility of registered reports
in the field of psychology, 
where still only 37 out of 62 studies supplied the code 
to redo the analysis \cite{obels2020analysis}.

@@

Genetic epidemiology is a field within biology that, among 
others, measures the spread of heritable traits,
as well as how these come to be.
For example, we know that lactose intolerance is, among others,
caused by producing too few lactose-degrading enzymes,
and is most commonly found in south-east asia.
The trait is caused by the genetic make-up, 
or genotype 
(see Table \ref{tab:definitions} for the definitions of terms used in this paper), 
of a person.
The trait, also called phenotype (again, see Table \ref{tab:definitions}), 
in this example is lactose intolerance,
yet any human property, such as weight or height can be studied.
Genetic epidemiology studies the spread of genotypes and phenotypes,
as well as the relation between those two.
When an association between genotype and phenotype is found,
these associations are subsequently used to 
create gene panels, where the gene causing 
an association is measured specifically, to, among others,
detect people at risk for the associated phenotype.

@@

The conclusions drawn from this paper,
may end up in the clinic.
For the sake of having a clear (yet fictitious) example,
let's assume that a high level of IL6RA 
is associated with a disease that develops later in life,
yet is preventable by lifestyle changes.
Would this be the case, we can create a gene panel that specifically measures
SNP \verb|rs12133641|. 
If the gene panel shows an individual has two guanines, 
we know that this person is likelier to develop higher levels of
IL6RA and is likelier to benefit from the lifestyle changes.

@@

The code for this experiment is vital paradata.
Like all paradata, its omission would be a constraint in the
use of the conclusion reached by it:
we can use it to assess the trustworthiness of the finding,
for example, by running the code again.
Ideally, the code is comprehensive, comprehensively documented
and be friendly to computers \cite{huvila2022improving}.
Code being friendly to computer seems like an obvious 'yes',
yet this is naive, as commonly it is hard to run code in practice. 
% TODO: Link to correct paragraphs
For code to be called computer-friendly, it must be run on an independent entity, 
as is discussed in the paragraphs about hosting code and/or putting the code in a container.
% TODO: Link to correct paragraphs
Also note that there is a problem due to the sensitivity of the data,
which is discussed in paragraph.
Lastly, note that the researcher used data that was itself generated by code:
complex bioinformatics pipelines are used to obtain reliable DNA sequences 
from the biological sample.

@@

The code of a computational experiment must be published
for an experiment to reproducible.
It is however (too) common that the original data
and/or code are unavailable \cite{peng2021reproducible}.
One saddening example is described in \cite{haibe2020importance}, where
an algorithm that detects breast cancer from images better than a human expert,
was irreproducible.
In the case of genetic epidemiology, 
the original data usually cannot be published, 
due to the guaranteed privacy of the people having shared their characteristics.
The code of a computational experiment, however, \emph{can} be
distributed without such problems.
In academic journals, however, its is not code that is published, 
but an English description of what it does instead.

From a knowledge management's perspective,
it is the code that must be archived 
to conserve an experiment,
and not the
English description of what it does:
it is the code does the actual work.
Hence, from that perspective, emphasis should be
put more on the preservation of code, as it is the most important actor
in an experiment.

@@

\begin{verbatim}
t.test(subjects_with_aa, subjects_with_gg)
\end{verbatim}

The line of code is simplified (as the result of the T test is never stored),
but the code is correct:
\verb|t.test| is indeed the name of an R function to do an unpaired T-test
and it is reasonable to assume that this code, when glancing over the code
of an article, is correct.

Here, however, we see a mismatch between the English description and the code:
by default, the R function \verb|t.test| does a \textbf{two}-tailed T-test.
A consequence of this fictional example is that the published p-values are
higher than needed, resulting in less significant findings, which
results in (needlessly) less conclusions drawn in this fictional paper.

@@

But regardless of the size of the code, 
it is the code that is the ground truth
and should be treated as a first-class citizen
in a computational experiment.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Making code useful paradata}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Again, paradata should be comprehensive, comprehensively documented
and friendly to computers \cite{huvila2022improving}.
In this section, these ideal properties of paradata are applied to code.

@@

@@

% Code must be reviewed
As it is the code that is the most important actor in an the experiment,
it should be treated as equally or more important than the scientific
paper describing what is does.
Regardless of this, most academic journals, 
when concerning computational experiments
only review the (usually) English description of the code
and make the review of code optional.
One notable exception is rOpenSci \cite{ram2018community},
which is an organisation that does peer-review academic code 
and facilitates the optional publication of a paper.
Although both code and paper are peer-reviewed,
most effort is spent on reviewing the code,
as can be seen in, for example, 
the \verb|mcbette| R package \cite{bilderbeek2020mcbette}.

% Code must be clean
% \paragraph{Raw code is an indicator of quality}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Code must be computer friendly}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% \paragraph{Raw code is already better than no code}

Publishing raw code is already better than not publishing code at all,
as that raw code can reasonably be assumed to have run successfully at least once.
To reproduce an experiment from raw code can be painful, 
yet it is usually better than writing the same code from scratch.
In general, it appears that sharing code is 
rare \cite{stodden2011trust,read2015sizing},
with a pleasant exception being \cite{conesa2019making}.

% \paragraph{Raw code can be FAIR data}

It is hard to make raw code be FAIR data. It is hard for computers to find
copy-pasted code in a Supplementary Materials. 
That code is usually only accessible
if the paper has been published under an open access license.
When code can be found and accessed, it is an open question if the
code can \emph{still} interoperate with current datasets, software
and hardware. Raw code usually lacks the metadata to describe its needs.
One could say that raw code, albeit better than no code, fits
FAIR data only poorly.

@@

% \paragraph{Example of hosted code}

@@

% \paragraph{Hosted code has important metadata}

@@

% \paragraph{Hosted code has a commit history}

@@

% \paragraph{A commit history helps assess honest reporting}

@@

% \paragraph{Continuous integration helps assess code quality}

@@

@@

Code is paradata, as it directly describes how the results are generated.
The part of a scholarly article 
that describes the experiment in a natural language is metadata, 
as it describes indirectly what the code does (or should be doing).
The code that does the work in a computational experiment should
always be published together with a scholarly article,
as, if the paper and code differ, it is the code that generated the results.

>>>>

\setcounter{chapter}{7}
\chapter{Research code is paradata and as vital as the article itself}

@@

Here we define paradata as data about the data collection process.
Computer code is the paradata for a computatioal/academic experiment,
allowing the experiment to be reproduced.
However, code is published alongside an academic
paper in a minority of all cases.
As reproducibility is a cornerstone of science,
research code should be published.
Earlier general recommendations about making paradata (more) useful
is tailored to code, using the field of genetic epidemiology
as an example.
The chapter concludes by some rules how to better code to serve as paradata,
and hence allowing computational research to be truly reproducible.

@@

% Alex and Blake?
Two different researchers in genetic epideomology (more on that field later), 
write two equally good manuscripts
that involve computation.
Both manuscripts are accepted by an equally prestigous journal after peer view. 
One researcher, however, does not supply the
computer code (from now on: 'code') that was used to generate the results,
where the other does.
Are the conclusion of these papers to be trusted equally?
Is this difference relevant and worth the effort?
This chapter discusses the effect of supplying the computer 
code  
and claims that code is paradata and that this paradata 
is just as important as the paper itself.

@@

A recent paper explores the use of paradata to increase the impact of data,
stating that lack of paradata can be seen as 'a drastic constraint'
in the use of data and offer some suggestions to 
make paradata useful for data re(use):
paradata should be comprehensive, comprehensively documented
and friendly to computers \cite{huvila2022improving}.
These suggestions appear to work with code as well,
where these suggestions show up as best practices (such
as \cite{stodden2013best}), some of which
described below.
The lack of computer code is just as much 'a drastic constraint'
in the use of a (computational) scientific experiment.
The clearest effect of its absence is, most obviously,
that a computational experiment cannot be redone 
easily, yet there are wider implications, described below, as well.

@@

This paper uses genetic epidemiology as a specific example,
to illustrate that the code is paradata,
that this type of paradata is relevant, as it helps 
to assess the correctness of the results,
where the correctness of the results is relevant for 
science in general and healthcare in particular.
However, any field that uses computation in its experiments
could be used as an example.

@@

Going back to the example scientists in the introduction,
where one does and the other does not publish the code,
it is not common to publish the code of an experiment or analysis 
\cite{stodden2011trust,read2015sizing} (with a pleasant exception 
being \cite{conesa2019making}).
For example, in computer graphics, 
a field intimately familiar with computer code,
only 192 of 454 SIGGRAPH papers supply computer code \cite{bonneel2020code}.
Another study analysed the reproducibility of registered reports
in the field of psychology, 
where still only 37 out of 62 studies supplied the code 
to redo the analysis \cite{obels2020analysis}.

@@

Genetic epidemiology is a field within biology that, among 
others, measures the spread of heritable traits,
as well as how these come to be.
For example, we know that lactose intolerance in adults is
caused by a decline in the production of lactose-degrading enzymes,
and is most commonly found in south-east asia and south africa \cite{storhaug2017country}.
The trait is caused by the genetic make-up, 
or genotype 
(see Table \ref{tab:definitions} for the definitions of terms used in this paper), 
of a person.
The trait, also called phenotype (again, see Table \ref{tab:definitions}), 
in this example is lactose intolerance at adult age,
yet any human property, such as weight or height can be studied.
Genetic epidemiology studies the spread of genotypes and phenotypes,
as well as the relation between those two.
When an association between genotype and phenotype is found,
these associations are subsequently used to 
create gene panels, where the gene causing 
an association is measured specifically, to, among others,
detect people at risk for the associated phenotype.

@@

The conclusions drawn from this paper,
may end up in the clinic.
For the sake of having a clear (yet fictitious) example,
let's assume that a high level of IL6RA 
is associated with a disease that develops later in life,
yet is preventable by lifestyle changes.
Would this be the case, we can create a tailored 
experiment, called a gene panel, that specifically measures
SNP \verb|rs12133641|. 
If the gene panel shows an individual has two guanines, 
we know that this person is likelier to develop higher levels of
IL6RA and is likelier to benefit from the lifestyle changes.

@@

The code for this experiment is vital paradata.
Like all paradata, its omission would be a constraint in the
use of the conclusion reached by it:
we can use it to assess the trustworthiness of the finding,
for example, by running the code again.
Ideally, the code is comprehensive, comprehensively documented
and be friendly to computers \cite{huvila2022improving}.
Code being friendly to computer seems like an obvious 'yes',
yet this is naive, as commonly it is hard to run code in practice. 
% TODO: Link to correct paragraphs
For code to be called computer-friendly, it must be run on an independent entity, 
as is discussed in the paragraphs about hosting code and/or putting the code in a container.
% TODO: Link to correct paragraphs
Also note that there is a problem due to the inherent sensitivity of the genetic data,
which is discussed in paragraph.
Lastly, note that the researcher used data that was itself generated by code:
complex bioinformatics pipelines are used to obtain reliable DNA sequences 
from the biological sample.

@@

The code of a computational experiment must be published
for an experiment to be easily reproducible.
It is however (too) common that the original data
and/or code are unavailable \cite{peng2021reproducible}.
One saddening example is described in \cite{haibe2020importance}, where
an algorithm that detects breast cancer from images better than a human expert,
was irreproducible.
In the case of genetic epidemiology, 
the original data usually cannot be published, 
due to the guaranteed privacy of the people having shared their characteristics.
The code of a computational experiment, however, \emph{can} be
distributed without such problems.
In academic journals, however, its is not code that is published, 
but an English description of what it does instead.

@@

\begin{verbatim}
t_test_result <- t.test(subjects_with_aa, subjects_with_gg)
\end{verbatim}

The code is correct:
\verb|t.test| is indeed the name of an R function to do an unpaired T-test
and it is reasonable to assume that this code, when glancing over the code
of an article, is correct.

Here, however, we see a mismatch between the English description and the code:
by default, the R function \verb|t.test| does 
a \textbf{two}-tailed unpaired T-test.
A consequence of this fictional example is that the published p-values are
higher than needed, resulting in less significant findings, which
results in (needlessly) less conclusions drawn in this fictional paper.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Making code useful paradata}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Again, paradata should be comprehensive, comprehensively documented
and friendly to computers \cite{huvila2022improving}.
In this section, these ideal properties of paradata are applied to code.
The resulting recommendations are compared with
existing best practices in writing code.

@@

For code to be ideal paradata, it must be documented.
However, code does explain itself to some extent,
for example, the function name \verb|t.test| as used above.

\begin{itemize}
  \item Code must be documented to the right amount
  \item Code must be reviewed
\end{itemize}

\paragraph{Code must be documented to the right amount}

Code should be documented to the right amount:
too little documentation is unhelpful, where too much is a needless burden
on the programmer.
The first recommended practice is to write the code in such a way 
that it becomes self-explanatory \cite{wilson2014best}.
What is left to document are the reasons behind the code, 
its design and its purpose \cite{wilson2014best}.

\paragraph{Code must be reviewed}

As it is the code that is the most important actor in an the experiment,
it should be treated as equally or more important than the scientific
paper describing what is does.
Regardless of this, most academic journals, 
when concerning computational experiments
only review the (usually) English description of the code
and make the review of code optional.
One notable exception is rOpenSci \cite{ram2018community},
which is an organisation that does peer-review academic code 
and facilitates the optional publication of a paper.
Although both code and paper are peer-reviewed,
most effort is spent on reviewing the code,
as can be seen in, for example, 
the \verb|mcbette| R package \cite{bilderbeek2020mcbette}.
Within software development best-practices, 
code reviews are recommended as well \cite{wilson2014best}

@@

For code to be ideal paradata, it must be extensive.
As code has many properties, there are many recommendations on this aspect.

\begin{itemize}
  \item Code must be clean
  \item Code must be hosted
  \begin{itemize}
    \item Hosted code generated important metadata
    \item Code must be tested
    \item ...
  \end{itemize}
  \item Code must be tested
  \begin{itemize}
    \item ...
  \end{itemize}
\end{itemize}

\paragraph{Code must be clean}

@@

\paragraph{Code must be hosted}

@@

\paragraph{Example of hosted code}

@@

\paragraph{Hosted code has important metadata}

@@

\paragraph{Hosted code keeps a history of changes}

@@

\paragraph{A commit history helps assess honest reporting}

@@

\paragraph{Code must be tested}

\paragraph{Continuous integration helps assess code quality}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Code must be computer friendly}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

This paper started with some suggestions to 
make paradata useful for data re(use):
paradata should be comprehensive, comprehensively documented
and friendly to computers \cite{huvila2022improving}.
When applying these general recomendations to code, 
this list can be phrased more precisely:
code should be comprehensive in supplying 
automatically generated metadata (such as commit history and code coverage).
The documentation should be as extensive as recommended by the 
software development literature. To be computer friendly,
code should be able to run on at least one independent entity,
such as a GitHub Actions server or from a Singularity container.
